---
title: "Project 2"
author: "Autumn Hakes"
date: "2024-03-13"
output: pdf_document
---

-continuously for 1 hour
-not large data set
-3 is interaction
-not going to be able to answer these with all one model
-just stationary in it? build from there
-build up a model simple -> not as simple
-there is correlation in the data - same kid measured every minute in an hour and there's 100 different kids
-For every kid, measured every minute for an hour – 59-60 measurements – a general symmetric correlation matrix would be 60*59/2… too many to handle – don’t use general symmetric – use something else
-questions build on themselves
-1 - only use stationary, find out how well this does
-2 - throw in activities
-3 - interactions between activities and kids - effect on activity kids specific - throw that in the model - is it better than the previous two
-4 - B0 + B1Activity1 + B2Activity1:Kid1 + B3Activity:Kid2
- (B1 + B2) and (B1 + B3) ... calculate them all and draw a histogram for every activity - use names of the coefficients
-use RMSE
-stationary needs to be in the final model even if it is not significant
-validating model assumptions - it might be beneficial if you consider transformations of the response and stationary - use the log scale


```{r}
library(tidyverse)
library(readr)
library(ggplot2)
library(forecast)
library(multcomp)
library(nlme)
library(corrplot)
library(mvtnorm)
library(car)
library(GGally)
source("stdres.gls.R")


data <- read_table("BreathingZonePM.txt", show_col_types = FALSE)
data$Activity <- as.factor(data$Activity)
```


```{r}
# Boxplot by Activity
ggplot(data, aes(x = Activity, y = Aerosol)) +
  geom_boxplot() +
  labs(title = "PM Measurements by Activity",
       x = "Activity",
       y = "PM Measurements on Vest")

# Scatterplot of Vest vs. Stationary
ggplot(data, aes(x = Stationary, y = Aerosol)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "PM Measurements on Vest vs. Stationary Monitor",
       x = "PM Measurements on Stationary Monitor",
       y = "PM Measurements on Vest")

# Scatterplot of Vest over Time Colored by Activity with Trend Lines
ggplot(data, aes(x = Minute, y = Aerosol, color = Activity)) +
  geom_point() +
  labs(title = "PM Measurements on Vest Over Time",
       x = "Minute",
       y = "PM Measurements on Vest")
```



```{r}
# Fit the linear model
lm_model <- lm(Aerosol ~ Stationary + Activity + Minute, data = data)

# Explore the residuals
residuals <- resid(lm_model)

# Organize residuals into a matrix and print it
residuals_matrix <- matrix(residuals, nrow=nrow(data), ncol = 3, byrow = TRUE)

cor(residuals_matrix)
```

There is high correlation.

## 1.

```{r}
model_ar1 <- gls(log(Aerosol) ~ log(Stationary),
                correlation = corAR1(form = ~ Minute | ID), method = "ML", data = data)

model_ma1 <- gls(log(Aerosol) ~ log(Stationary),
                correlation = corARMA(form = ~ Minute | ID, p = 1, q = 1),
                data = data)

AIC(model_ar1)
AIC(model_ma1)
summary(model_ar1)
```

With just stationary, AR1 is lower AIC = better

```{r}
# Predict the values using the model
predicted_values <- predict(model_ar1)
observed_values <- data$Aerosol

# Calculate the RMSE
rmse <- sqrt(mean((observed_values - predicted_values)^2))
print(rmse)

# Calculate Pseudo R-squared
pseudo_r_squared <- cor(observed_values, predicted_values)^2
print(pseudo_r_squared)
```
RMSE = 10.40912: This value indicates that, on average, the model's predictions are about 10.40912 units away from the actual 'Aerosol' values. In context of the data is 10.4 lower or higher a large value?

Pseudo R-squared = 0.00313318: This value is very close to 0, suggesting that the model explains very little of the variance in the 'Aerosol' data — only about 0.31%. A low value like this one implies that the independent variables included in the model have very little predictive power concerning the dependent variable.

In summary, the RMSE is relatively high (considering the context of 'Aerosol' values), and the Pseudo R-squared is very low. Both of these suggest that the model may not be performing well in predicting the 'Aerosol' variable. It's worth considering model diagnostics and potentially looking into model improvements, such as feature engineering, model selection, or addressing any data quality issues. It could also be informative to compare this model to simpler or more complex models, or even to different types of models, to see if they can provide better predictive accuracy.

## 2. 
```{r}
model_ar2 <- gls(log(Aerosol) ~ log(Stationary) + Activity,
                correlation = corAR1(form = ~ Minute | ID), method = "ML", data = data)

model_ma2 <- gls(log(Aerosol) ~ log(Stationary) + Activity,
                correlation = corARMA(form = ~ Minute | ID, p = 1, q = 1),
                data = data)

AIC(model_ar2)
AIC(model_ma2)

anova(model_ar1, model_ar2)
```

Adding Activity, MA2 is better. It is also lower AIC than MA1.

```{r}
model_ar3 <- gls(log(Aerosol) ~ log(Stationary) + Activity + Activity:ID,
                correlation = corAR1(form = ~ Minute | ID), method = "ML", data = data)

model_ma3 <- gls(log(Aerosol) ~ log(Stationary) + Activity + Activity:ID,
                correlation = corARMA(form = ~ Minute | ID, p = 1, q = 1),
                data = data)

AIC(model_ar3)
AIC(model_ma3)

summary(model_ar3)
```

Adding interaction between activity and kid, AR3 has a lower AIC and it's lower than MA1 and MA2.

```{r}
# Linear Assumption
avPlots(lm_model)


# Check independence using the decorrelated residuals
residuals <- stdres.gls(model_ar3)

# Organize residuals into a matrix and print it
residuals_matrix2 <- matrix(residuals, nrow=nrow(data), ncol = 3, byrow = TRUE)

cor(residuals_matrix2)


# Check normality using histogram of decorrelated residuals
ggplot() +
  geom_histogram(mapping = aes(x=residuals))


# Check equal variance using a fitted vs decorrelated residuals plot
ggplot()+
  geom_point(mapping = aes(x=fitted(model_ar3), y = residuals))
```

