---
title: "Project 2"
author: "Autumn Hakes"
date: "2024-03-13"
output: pdf_document
---

-continuously for 1 hour
-not large data set
-3 is interaction
-not going to be able to answer these with all one model
-just stationary in it? build from there
-build up a model simple -> not as simple
-there is correlation in the data - same kid measured every minute in an hour and there's 100 different kids
-For every kid, measured every minute for an hour – 59-60 measurements – a general symmetric correlation matrix would be 60*59/2… too many to handle – don’t use general symmetric – use something else
-questions build on themselves
-1 - only use stationary, find out how well this does
-2 - throw in activities
-3 - interactions between activities and kids - effect on activity kids specific - throw that in the model - is it better than the previous two
-4 - B0 + B1Activity1 + B2Activity1:Kid1 + B3Activity:Kid2
- (B1 + B2) and (B1 + B3) ... calculate them all and draw a histogram for every activity - use names of the coefficients
-use RMSE
-stationary needs to be in the final model even if it is not significant
-validating model assumptions - it might be beneficial if you consider transformations of the response and stationary - use the log scale


```{r}
library(tidyverse)
library(readr)
library(ggplot2)
library(forecast)
library(multcomp)
library(nlme)
library(corrplot)
library(mvtnorm)
library(car)
library(GGally)
source("stdres.gls.R")


data <- read_table("BreathingZonePM.txt", show_col_types = FALSE)
data$Activity <- as.factor(data$Activity)
```


```{r}
# Boxplot by Activity
ggplot(data, aes(x = Activity, y = Aerosol)) +
  geom_boxplot() +
  labs(title = "PM Measurements by Activity",
       x = "Activity",
       y = "PM Measurements on Vest")

# Scatterplot of Vest vs. Stationary
ggplot(data, aes(x = Stationary, y = Aerosol)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "PM Measurements on Vest vs. Stationary Monitor",
       x = "PM Measurements on Stationary Monitor",
       y = "PM Measurements on Vest")

# Scatterplot of Vest over Time Colored by Activity with Trend Lines
ggplot(data, aes(x = Minute, y = Aerosol, color = Activity)) +
  geom_point() +
  labs(title = "PM Measurements on Vest Over Time",
       x = "Minute",
       y = "PM Measurements on Vest")
```



```{r}
# Fit the linear model
lm_model <- lm(Aerosol ~ Stationary + Activity + Minute, data = data)

# Explore the residuals
residuals <- resid(lm_model)

# Organize residuals into a matrix and print it
residuals_matrix <- matrix(residuals, nrow=nrow(data), ncol = 3, byrow = TRUE)

cor(residuals_matrix)
```

There is high correlation.

```{r}
model_ar1 <- gls(log(Aerosol) ~ log(Stationary),
                correlation = corAR1(form = ~ Minute | ID), method = "ML", data = data)

model_ma1 <- gls(log(Aerosol) ~ log(Stationary),
                correlation = corARMA(form = ~ Minute | ID, p = 1, q = 1),
                data = data)

AIC(model_ar1)
AIC(model_ma1)
```

With just stationary, MA1 is lower AIC = better

```{r}
model_ar2 <- gls(log(Aerosol) ~ log(Stationary) + Activity,
                correlation = corAR1(form = ~ Minute | ID), method = "ML", data = data)

model_ma2 <- gls(log(Aerosol) ~ log(Stationary) + Activity,
                correlation = corARMA(form = ~ Minute | ID, p = 1, q = 1),
                data = data)

AIC(model_ar2)
AIC(model_ma2)
```

Adding Activity, MA2 is better. It is also lower AIC than MA1.

```{r}
model_ar3 <- gls(log(Aerosol) ~ log(Stationary) + Activity + Activity:ID,
                correlation = corAR1(form = ~ Minute | ID), method = "ML", data = data)

model_ma3 <- gls(log(Aerosol) ~ log(Stationary) + Activity + Activity:ID,
                correlation = corARMA(form = ~ Minute | ID, p = 1, q = 1),
                data = data)

AIC(model_ar3)
AIC(model_ma3)
```

Adding interaction between activity and kid, AR3 has a lower AIC and it's lower than MA1 and MA2.

```{r}
# Linear Assumption
avPlots(lm_model)


# Check independence using the decorrelated residuals
residuals <- stdres.gls(model_ar3)

# Organize residuals into a matrix and print it
residuals_matrix2 <- matrix(residuals, nrow=nrow(data), ncol = 3, byrow = TRUE)

cor(residuals_matrix2)


# Check normality using histogram of decorrelated residuals
ggplot() +
  geom_histogram(mapping = aes(x=residuals))


# Check equal variance using a fitted vs decorrelated residuals plot
ggplot()+
  geom_point(mapping = aes(x=fitted(model_ar3), y = residuals))
```

